---
title: "Coffee Sales Overtime"
author: "Terence Morita"
date: "2024-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
data <- read_csv("data_2024_11_06.csv")
```

## Introduction

The objective of this analysis is to determine coffee flavour popularity, and its relation to the time of the year. The data set contains data between March 2024 and October 2024 inclusive. This is a relatively new data set and does not contain the months between November and February inclusive. So, I am planning to reanalyze the dataset once the data from missing months become available. The analysis works with an [open source dataset](https://www.kaggle.com/datasets/ihelon/coffee-sales/data) from Kaggle, by Yaroslav Isaikenkov. 

## Cleaning Data

To ensure data is clean to work with, I checked for duplicates, missing or invalid values, and misspellings.*Note that card column should be empty if payment method is cash*

```{r}
## View quick summary on data

summary(data)
str(data)

## Ensure there are no mispelled coffee names
unique(data$coffee_name)

## Check for unusual coffee prices
unique(data$money)

## Remove duplicate data
clean_data <- data %>% distinct()

## Remove data with empty entries, with the exception of "card" when "cash_type" == "cash". 

clean_data <- clean_data %>%
  filter(
    !(is.na(card) & cash_type != "cash") &  
    !if_any(everything(), is.na)           
  )

```
## Price Changes Over the Year

Price is an important factor customers consider when making purchases. Running the code below will provide a visualization of the price changes over the year for each coffee flavour. Note that every flavour undergoes the same transformation of price change over the year. Also note that the most drastic price changes occur between June and July, and September to October.

```{r} 

# Convert date time column
clean_data <- clean_data %>%
  mutate(month = month(date, label = TRUE, abbr = TRUE))



# Calculate the average price for each coffee flavour by month
price_data <- clean_data %>%
  group_by(month, coffee_name) %>%
  summarise(avg_price = mean(money, na.rm = TRUE)) %>%
  arrange(match(month, c("Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")))  # Order months correctly

# Create the plot
price_plot <- ggplot(price_data, aes(x = month, y = avg_price, color = coffee_name, group = coffee_name)) +
  geom_line() +  # Plot the price change as lines
  geom_point() +  # Add points to make the graph clearer
  labs(title = "Price Changes of Coffee Flavours (March to October 2024)",
       x = "Month",
       y = "Average Price",
       color = "Coffee flavour") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

# Print the price change plot
print(price_plot)
```




## Visualization

We now visualize the proportions of coffee flavour purchases for every month between March 2024 to October 2024. 

```{r}

months_visual <- c("Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct") ## Months contained in dataset


## Iterate code for each month
for (month_name in months_visual) {
  
 
  month_data <- clean_data %>%
    filter(month == month_name)  
  
  ## Indicate variables of interest
  coffee_by_month <- month_data %>%
    group_by(coffee_name) %>%
    summarise(purchase_count = n()) %>%
    arrange(desc(purchase_count)) %>%
    mutate(percentage = purchase_count / sum(purchase_count) * 100)  
  
  ## Visualize data as pie charts
  pie_chart <- ggplot(coffee_by_month, aes(x = "", y = purchase_count, fill = coffee_name)) +
    geom_bar(stat = "identity", width = 1) +    
    coord_polar(theta = "y") +                 
    labs(title = paste("Coffee Type Popularity in", month_name)) +  
    theme_void() +                             
    theme(legend.title = element_blank()) +      
    geom_text(aes(label = paste0(round(percentage, 1), "%")),   
              position = position_stack(vjust = 0.5),           
              color = "white")  

  print(pie_chart)
}
 
```

## Analysis

**Important note that unfortunately, the dataset does not contain data between Nov and Feb. I will reanalyze the dataset when data from missing months become available. For now, I will anaylze the current available data. (2024-11-06)**

**Following observations are made per flavour, organized by price bracket:**

**Cappuccino** spikes in popularity in April, which extends until the end of June. In the other months, it remains at a fairly consistent purchase rate. This may indicate that cappuccino is a preferred flavour during mid to late Spring. 

**Cocoa** is very unpopular throughout the year. However, we could see that it relatively increases in Oct. This may indicate that winter favours cocoa, however the winter data does not exist to confirm this.

**Hot Chocolate** spikes in popularity during October, which may indicate that customers enjoy Hot Chocolate during colder seasons. Also, March has a higher Hot Chocolate count compared to the months after. We may be able to deduce that Hot Chocolate purchase gets higher between Oct and March as a result. Additional data from November to February could confirm this. 

**Latte** remains a popular choice all year, with a lower than usual purchase during specifically April. It is even more popular during the fall months. 

------------------------------------------------------------------------------

**Americano with Milk** spikes in popularity as summer approaches, and remains popular (relative to other months) until October. Seeing that it is relatively not as popular during March, this may imply that this flavour gets less popular during the cold months of Oct to Feb. Unfortunately, the lack of winter data makes it difficult to confirm this. Despite Americano being considerably cheaper than its milk counterpart, it is still consistently less popular. Most people are willing to pay additional money for the milk. 

------------------------------------------------------------------------------

**Americano** is consistently a less popular choice than its milk containing counterpart. March is the only month in the dataset which shows milk-less Americano to be more popular. As a result, Winter season may affect this preference which extends to March, but the current dataset cannot confirm this. 

**Cortado** does not appear to follow any seasonal trend in the given months.

------------------------------------------------------------------------------

**Espresso** is an unpopular choice all year round. 









